---
title: "SAT and College Grades"
author: "Daniel Heitz"
date: 2023-04-19
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
---

*Purpose*: How do we apply hypothesis testing to investigating data? In this challenge you'll practice using hypothesis testing tools to make sense of a dataset.

*Reading*:
- [Harvard Study Says SATs Should Be Optional: Here's Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why) (Optional); easy-to-read news article on colleges going SAT-free
- [Norm-Referenced Tests and Race-Blind Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university) (Optional); technical report on relationship between the SAT/ACT and non-academic factors

*Credit*: This is based on a [case study](http://onlinestatbook.com/2/case_studies/sat.html) originally prepared by Emily Zitek, with data collected through the research of Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
```

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

*Background*: Every year about 2 million students take the Scholastic Aptitude Test (SAT). The exam is [controversial](http://www.nea.org/home/73288.htm) but [extremely consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html). There are many claims about the SAT, but we're going to look at just one: Is the SAT predictive of scholastic performance in college? It turns out this is a fairly complicated question to assess---we'll get an introduction to some of the complexities.

# Obtain the Data
<!-- -------------------------------------------------- -->

### __q1__ Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
df_sat <- read_excel("sat.xls")
df_sat
summary(df_sat)
```

**Observations**:

- Fill in the following "data dictionary"

| Column     | Meaning |
|------------|---------|
| `high_GPA` | ???     |
| `math_SAT` | ???     |
| `verb_SAT` | ???     |
| `comp_GPA` | ???     |
| `univ_GPA` | ???     |

- What information do we have about these students?
  - (Your response here)
- What kinds of information *do we not have* about these students?
  - (Your response here)
- Based on these missing variables, what possible effects could be present in the data that we would have *no way of detecting*?
  - (Your response here)

# Analysis with Hypothesis Testing
<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the first based on hypothesis testing of correlation coefficients, and the second based on fitting a regression model and interpreting the regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)
```

## View 1: Correlations
<!-- ----------------------------------------------------------------------- -->

### __q2__ Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`.

```{r q2-task}
## TODO:
df_composite %>% 
  gather(key = "facet", value = "y", high_GPA, both_SAT) %>%
  ggplot(aes(x = univ_GPA, y = y)) +
  geom_point() +
  # Facet by the gathered column
  facet_wrap(~facet, scales = "free_y")
```


**Observations**:

- What relationship do `univ_GPA` and `both_SAT` exhibit?
  - Positive, accelerating correlation
- What relationship do `univ_GPA` and `high_GPA` exhibit?
  - Positive, accelerating correlation

### Hypothesis Testing with a Correlation Coefficient
<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation coefficient. The
idea is to set our null hypothesis to the case where there is no correlation,
and test to see if the data contradict that perspective. Formally, the null (H0)
and alternative (HA) hypotheses relating to a correlation coefficient between
two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the assumption that `X, Y` are both normally distributed. First, let's check to see if this assumption looks reasonable for our data.

### __q3__ Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}
library(gridExtra)

# Create the first histogram for univ_GPA
p1 <- df_composite %>%
  ggplot(aes(x = univ_GPA)) +
  geom_histogram(binwidth = 0.2, fill = "red") +
  ggtitle("univ_GPA")

# Create the second histogram for high_GPA
p2 <- df_composite %>%
  ggplot(aes(x = high_GPA)) +
  geom_histogram(binwidth = 0.2, fill = "green") +
  ggtitle("high_GPA")

# Create the third histogram for both_SAT
p3 <- df_composite %>%
  ggplot(aes(x = both_SAT)) +
  geom_histogram(binwidth = 30, fill = "blue") +
  ggtitle("both_SAT")

# Arrange the three histograms next to each other
grid.arrange(p1, p2, p3, ncol = 3)
```

**Observations**:

- To what extent does `both_SAT` look like a normal distribution?
  - Its not great but its pretty good. I played around with the binwidth and in all of them it bore some resemblance so I'd say it probably is.
- To what extent does `high_GPA` look like a normal distribution?
  - Pretty good, there's a hole in the middle but on both sides it looks fine, so I'd go with a fairly strong resemblance
- To what extent does `univ_GPA` look like a normal distribution?
  - It actually looks like two of them, one smaller than the other. Overall not very, but each peak definitely looks somewhat like one (though quite steep)

Keep in mind your findings as you complete q4.

### __q4__ Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
##       both_SAT and univ_GPA
# Calculate the correlation and confidence interval between high_GPA and univ_GPA
high_univ_cor <- cor.test(df_composite$high_GPA, df_composite$univ_GPA)
high_univ_df <- tibble(
  estimate = high_univ_cor$estimate,
  conf.low = high_univ_cor$conf.int[1],
  conf.high = high_univ_cor$conf.int[2]
) %>%
  summarise(
    "Correlation between high_GPA and univ_GPA" = estimate,
    "Lower bound of confidence interval" = conf.low,
    "Upper bound of confidence interval" = conf.high
  )

# Calculate the correlation and confidence interval between both_SAT and univ_GPA
both_univ_cor <- cor.test(df_composite$both_SAT, df_composite$univ_GPA)
both_univ_df <- tibble(
  estimate = both_univ_cor$estimate,
  conf.low = both_univ_cor$conf.int[1],
  conf.high = both_univ_cor$conf.int[2]
) %>%
  summarise(
    "Correlation between both_SAT and univ_GPA" = estimate,
    "Lower bound of confidence interval" = conf.low,
    "Upper bound of confidence interval" = conf.high
  )
high_univ_df
both_univ_df
```

**Observations**:

- Which correlations are significantly nonzero?
  - (Your response here)
- Which of `high_GPA` and `both_SAT` seems to be more strongly correlated with `univ_GPA`?
  - high_GPA seems to be more strongly correlated, although the interval has enough overlap that I cannot say this definitively.
- How do the results here compare with the visual you created in q2?
  - It about lines up, yeah they all look similar but the GPA graphs both have a double peak behavior with the double peaks being more distinct in univ_GPA, but each peak and overall ignoring the gap all resemble similar normal distributions.
- Based on these results, what can we say about the predictive capabilities of both `high_GPA` and `both_SAT` to predict `univ_GPA`?
  - They are both fairly good predictors, especially high_GPA. Its not amazing, but it is definitely a strong correlation that easily gives us a qualitative sense.

Finally, let's use the bootstrap to perform the same test using *different* assumptions.

### __q5__ Use the bootstrap to approximate a confidence interval for `corr[high_GPA, univ_GPA`. Compare your results---both the estimate and confidence interval---to your results from q4.

```{r q5-task}
## TODO: Use the bootstrap to compute a confidence interval for corr[high_GPA, univ_GPA]
# Set the number of bootstrap replicates
# Set the number of bootstrap replicates
n_replicates <- 1000

# Create a function to calculate the correlation for a given bootstrap replicate
calc_correlation <- function(data) {
  cor(data$high_GPA, data$univ_GPA)
}

# Generate bootstrap replicates of the correlation
correlation_replicates <- map_dbl(1:n_replicates, ~ df_composite %>%
  sample_frac(replace = TRUE) %>%
  calc_correlation())

# Calculate the estimate and confidence interval for the correlation
correlation_estimate <- mean(correlation_replicates)
correlation_ci <- quantile(correlation_replicates, c(0.005, 0.995))

# Display the results
tibble(
  estimate = correlation_estimate,
  conf.low = correlation_ci[1],
  conf.high = correlation_ci[2]
) %>%
  summarise(
    "Bootstrap estimate of correlation between high_GPA and univ_GPA" = estimate,
    "Lower bound of confidence interval" = conf.low,
    "Upper bound of confidence interval" = conf.high
  )
```

**Observations**:

- How does your estimate from q5 compare with your estimate from q4?
  - Its a little bigger but it does fully include the old estimate.
- How does your CI from q5 compare with your CI from q4?
  - They're almost exactly the same, 0.780 vs 0.781.

*Aside*: When you use two different approximations to compute the same quantity and get similar results, that's an *encouraging sign*. Such an outcome lends a bit more credibility to the results.

## View 2: Modeling
<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study the relationship among more variables we can instead use a fitted model. Using a model, we can also help assess whether it is *worthwhile* to measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)

df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)

df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model
<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a model, we can express our hypotheses in terms of the model parameters. For instance, if we were interested in whether $X$ has an affect on $Y$, we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on $Y$. Let's apply this idea to relating the variables `univ_GPA` and `high_GPA`. Luckily R has built-in tools to construct a confidence interval on the $\beta$'s in a regression [1]; we'll simply use those tools rather than do it by hand.

### __q6__ Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <- lm(univ_GPA ~ both_SAT, data = df_composite)

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

model_summary <- summary(fit_basic)

# Get the R-squared value
r_squared <- model_summary$r.squared

# Print the R-squared value
r_squared
```

**Observations**:

- What is the confidence interval on the coefficient of `both_SAT`? Is this coefficient significantly different from zero?
  - 0.001984 and 0.00349. Yes I would say this is significantly different because the interval doesn't include 0.
- By itself, how well does `both_SAT` predict `univ_GPA`?
  - I would say fairly decently. Its not great, the R2 value is only 0.47. This definitely suggests a real relationship though. I would characterize our regression as a good ballpark estimate but not something we should follow religiously.

Remember from `e-model03-interp-warnings` that there are challenges with interpreting regression coefficients! Let's investigate that idea further.

### __q7__ Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone
# Fit the linear model with both_SAT and high_GPA as predictors
model1 <- lm(univ_GPA ~ both_SAT + high_GPA, data = df_composite)

# Summarize the model
summary(model1)

# Fit the linear model with only high_GPA as predictor
model2 <- lm(univ_GPA ~ high_GPA, data = df_composite)

# Summarize the model
summary(model2)

# Assess the performance of model1

# Get the summary of model1
model1_summary <- summary(model1)

# Get the R-squared value for model1
r_squared1 <- model1_summary$r.squared

# Print the R-squared value for model1
cat("R-squared value for model1:", r_squared1, "\n")

# Get the p-values for each predictor in model1
p_values1 <- model1_summary$coefficients[, 4]

# Print the p-values for model1
cat("p-values for predictors in model1:\n")
print(p_values1)

# Create a residual plot for model1
plot(model1, which = 1)

# Get the 99% confidence intervals for all coefficients in model1
ci_model1 <- confint(model1, level = 0.99)

# Print the 99% confidence intervals for all coefficients in model1
cat("99% confidence intervals for all coefficients in model1:\n")
print(ci_model1)

# Assess the performance of model2

# Get the summary of model2
model2_summary <- summary(model2)

# Get the R-squared value for model2
r_squared2 <- model2_summary$r.squared

# Print the R-squared value for model2
cat("R-squared value for model2:", r_squared2, "\n")

# Get the p-values for each predictor in model2
p_values2 <- model2_summary$coefficients[, 4]

# Print the p-values for model2
cat("p-values for predictors in model2:\n")
print(p_values2)

# Create a residual plot for model2
plot(model2, which = 1)

# Get the 99% confidence intervals for all coefficients in model2
ci_model2 <- confint(model2, level = 0.99)

# Print the 99% confidence intervals for all coefficients in model2
cat("99% confidence intervals for all coefficients in model2:\n")
print(ci_model2)
```

**Observations**:

- How well do these models perform, compared to the one you built in q6?
  - Even better, their R2 values are improved and both have great p-values.
- What is the confidence interval on the coefficient of `both_SAT` when including `high_GPA` as a predictor?? Is this coefficient significantly different from zero?
  - -0.00022 to 0.00181. This is not significantly different from zero.
- How do the hypothesis test results compare with the results in q6?
  - These results suggest that both_SAT is an inferior predictor than high_GPA, and actually suggests that both_SAT is not a good predictor in general which goes against what we found in part 6.

## Synthesize
<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### __q8__ Using the results from all previous q's, answer the following questions.

**Observations**:

- Between `both_SAT` and `high_GPA`, which single variable would you choose to predict `univ_GPA`? Why?
  - high_GPA, all the metrics favor it as a predictor of univ_GPA.
- Is `both_SAT` an effective predictor of `univ_GPA`? What specific pieces of evidence do you have in favor of `both_SAT` being effective? What specific pieces of evidence do you have against?
  - Yes I would say it is. Our bootstrap confidence intervals, the regression from q6, the bar graphs, and the dot plots all suggest there is some relationship. The evidence against is the findings in q7, but I more interpret that as saying that it is a significantly worse predictor than high_GPA, not that it isn't able to produce any reasonable prediciton.

# End Notes
<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more information see this [Wiki article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
